---
output:
  html_document:
    keep_md: yes
---




# 1. Standard Preprocessing using Seurat


Some standard steps are usually carried out in scRNA-Seq prior to further analysis, such as quality control (QC), dimensional
reduction and marker visualization. Here, we will use the Seurat R package to perform these steps which
is increasingly becoming the most popular tool. However, there are  other options such as scanpy which is available for python. First, we need to define a **Seurat object**.

The Seurat object is a special data structure, which contains multiple type of information: the raw counts, the feature names (here genes), the cell identifiers, if available some processed data and additional metadata information. As we process through the analysis, we will keep on filling the Seurat object with new pieces of information (variable genes, principle components, etc...)


## Creating Seurat object

We create a Seurat object using the `CreateSeuratObject` function as follows. Use the `?function` helper
in R to get information about the parameters that are need to be provided to the function.

 * counts - a count matrix. It can be a matrix, sparse matrix or dataframe.
 * project - A single character string. Correspond to a arbitrary name to label the object.
 * assay - A single character string. An arbitrary name that usually is assigned to label the
            type of sequencing information, for example, RNA, spliced RNA, ATAC, etc...
 * min.cells - An integer. Indicates a threshold of the number of cells for which a feature was
              recorded. Cells below that threshold will be filtered out.
 * min.features - An integer. Similar to min cells but for the number of features of a cell.
 


```{r, message=FALSE, warning=FALSE}
ad_medula_seurat <- CreateSeuratObject(
  counts = counts, 
  project = 'human_ad_medulla', 
  assay = 'RNA', 
  min.cells = 1, 
  min.features = 1
)
Idents(ad_medula_seurat) <- 'human_ad_medulla'
```

The variable `ad_medula_seurat` now contains the Seurat object that we can feed into the package.
If we print the variable we get information about the number of genes and cells.


```{r}
ad_medula_seurat
```



## Exploring the Seurat object


Seurat objects can be seen as a container of different features. At this step it contains
our gene expression matrix, but in addition it can store metadata, processed data,
information from different assays, for example, scATACSeq, scCITESeq or unspliced transcripts.

We can explore the seurat object using the `$` to explore its *metadata* in combination with the tab 
key. For example, during the creation of the seurat object the number of counts quality metric
is calculated and added to the metadata. We can explore this metric by accessing the metadata
as follows.


```{r, warning=FALSE, message=FALSE}
ad_medula_seurat$nCount_RNA %>% head
```


 We can do the same with the `@` operator to explore the different *slots*. For example,
 we can extract the original count matrix that we used to create the seurat object as follows:
 
 

```{r}
ad_medula_seurat@assays$RNA$counts[1:5, 1:5]
```


## Extracting expression values 


Next, we visualize gene counts to see its behavior. We take a look at the expression of the 
house keeping gene ACTIN Beta and plot an histogram of count values. We will use the
the function `FetchData` which is used to extract values from selected features in the Seurat
object and then plot it using an histogram.



```{r, message=FALSE, warning=FALSE}
actin <- FetchData(ad_medula_seurat, vars = 'ACTB')
hist(actin$ACTB)
```





## Quiz

<details>
<summary> How can you find and display the metadata in the seurat object?
<br>
<i>TIP: You can have 
a look at the <a href="https://github.com/satijalab/seurat/wiki/Seurat#object-information">documentation</a>
of the seurat objects from the GitHub Wiki.</i>
</summary>
<br>
<b>Answer:</b>
<br>

```r
ad_medula_seu_filtered@meta.data %>% head
```

</details> 



## Exercise

<blockquote>
Create a Seurat object 

The file in the follwing URL:

`https://raw.githubusercontent.com/caramirezal/caramirezal.github.io/master/bookdown-minimal/data/pbmc_10X_250_cells.tsv` 

contains 250 cells downsampled from the 10x PBMC data and stored in tsv format

 * Load the count matrix in tsv format
 
 * Create a Seurat object using the count matrix
 
 * How many features and cells are present in the count matrix?

</blockquote>




[Previous Chapter (Seurat)](./01-Seurat.md)|
[Next Chapter (Feature selection)](./03-Feature_selection.md)





