---
output: 
  html_document: 
    keep_md: yes
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print=200)
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=150)
```

# 2. First steps in the analysis of a data matrix

## 2.1 reading in the data!

Most of the time, the dataset you want to analyse comes in the form of a large data matrix, with columns and rows. Typically, an Excel sheet!

In this section, we will learn how to read in such a table, and start doing some first exploratory analysis on it!

The diabetes dataset, which we will be using throughout this practical will be downloaded from an online repository. We will load that into R and have a sneak peek into how it looks like with the console. In the following you will see several functions that give us information about our dataset. You will also see how the "\$" sign is used to access specific columns of out dataframe.

If your data file is in the form of a text file, with tab delimited columns (typically with the ending `.tsv` or `.csv`), you can use the following command to read the data file

```{r}
dat = read.delim('https://tinyurl.com/ex9hxvvr', stringsAsFactors = FALSE)
head(dat, 10) # Look at the first 10 lines of the table
```

**Note that we indicated here a url to a remote file! If you have instead a local file stored on your computer, indicate the path to this file instead of the url!**

<detail>

<summary>

<b>What if I have an Excel sheet to read in?</b>

</summary>

If on the other hand the file is an Excel file, then you can use a specific library containing a command to read in Excel formated files.

However, we cannot indicate a URL in the function; so we first need to download the file to a local folder, and then read the file from this local storage.

-   Download the file under [this link](https://tinyurl.com/25x5t6wr); note that this is an excel file with `.xlsx` ending. **Please remember where you downloaded the file, and the path to it!!**

-   Run the following code, and replace the path string with the path to the file on your computer!

```{r}
# we need to load the library xlsx first
library('xlsx')
# here, replace the path with the path to the file!
dat.xls = read.xlsx('~/Dropbox/IRTG2021/data/diabetes_full.xlsx',1,stringsAsFactors=FALSE)
```

However, the `read.xlsx` function might sometimes not recognize the data type of the columns correctly, and e.g. interpret a column of numerical values as strings...

An alternative could there for be:

1.  read in your excel sheet in the Excel program

2.  use "save as..." to export the sheet in tab-separated format

3.  read in using the `read.table` function </detail>

    <p>

    </p>

## 2.2 explore and understand our dataset

Now that we have read in the data, we can start analysing it! First of all, what is the data type of the data we just read?

```{r}
class(dat)
```

Yes, this is a data frame, the closest equivalent in R to an Excel sheet. It can contain all sorts of data (numbers, boolean, strings,...)

Let's start by asking some simple questions:

-   what is the dimension of our dataset? (i.e. how many rows/columns are there in our data)

```{r}
# Dimension
dim(dat)
```

```{r, results='hide'}
# Number of columns
ncol(dat)
# Number of rows
nrow(dat)
```

-   what are the column names of our dataset?

```{r}
colnames(dat) # Similarly rownames for rows
```

Probably you are confused about what these column names mean. For more description on these values [look here](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/Cdiabetes.html)

-   how do we extract the minimum and maximum age of patients in our dataset?

```{r, results='hide'}
min(dat$age)
max(dat$age)
range(dat$age)
# Try to find out the same for height and weight
```

-   how does the overall summary of our entire dataset look like?

```{r rows.print=10}
summary(dat) # Can you explain what you see?
```

*Note how the output of the summary function differs depending on the data type of the columns (numeric or strings)!*

You have just seen a lot of new functions. If you're wondering how any of them work, you can use the function "help()" or the sign "?" in the console to read more about any function in the bottom right pane. For example:

```{r eval = FALSE}
help("max")
?max
```

will both get you more information on this function. Remember that every function will automatically print its result in the console pane, unless it is assigned to a variable. You can try by running the following line of code:

```{r eval = FALSE}
max(dat$age)
MaxAge = max(dat$age)
```

See the difference? Now that value is saved into this new variable. Before we keep going, here is a quick reminder of how indexing and slicing works in R. If you're already familiar with it you can skip the next chunk.

<details>

<summary>

<b>HOWTO: access rows/columns/cells in a data frame </b>

</summary>

```{r eval = FALSE}

# Returning a specific column or line of a data structure
dat[1,] # Returns the first line
dat[,1] # Returns the first column
dat[1,1] # Returns only the first object of the first line


# Returning an interval of columns or lines of a data structure 
dat[1:3,] # Returns the first three lines
dat[,1:3] # Returns the first three columns
dat[1:3,1:3] # Returns the first three elements of the first three lines

# If your data frame has column names, you can use this to extract a column
dat$age# Returns the column age

# BEWARE!! This last method DOES NOT WORK for matrices (even if they have column names...)
```

Feel free to play around with this syntax until you feel comfortable with it. You can open a window with View(dat) to compare your results.

</details>

Let's do some practice with the dataset we have downloaded and the new functions we have learned.

#### Exercise 2.1: manipulating data frames

1.  Compute the mean age of the patients in the dataset. Remember that R functions often have intuitive names....
2.  Compute the mean hdl level for the patients; do you understand why there is a problem here?
3.  The weights (column `weight`) are given in pounds; convert the weights in this columns into kilograms, by multiplying the column by 2.25
4.  Same as question 3, but now assign the result to a new column in the data frame, called `weight.kg`.

<details>

<summary>

Click for solution!

</summary>

```{r}
## mean age
mean(dat$age)

## mean hdl
mean(dat$hdl)  # gives NA because there are NA values ...

mean(dat$hdl,na.rm=TRUE)

dat$weight.kg = dat$weight*2.25
```

</details>
