---
title: "Single Cell ATAC-Seq"
author: "Ramirez C and Herrmann C"
date: "2024-06-10"
output:
  html_document: 
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
        warning = FALSE, 
        message = FALSE,
        cache = FALSE,
        fig.align='center'
)
```

# 9. single-cell ATAC-seq from blood cells


In this tutorial we introduce an standard pipeline of Single Cell ATAC-Seq
analysis. In order to run the steps described below please download the
following files:

* [fragment files](https://drive.google.com/file/d/1CNkvJQeXwhE44vA_TvNL9rb-HDlFYjlM/view?usp=sharing)
* [index file](https://drive.google.com/file/d/1fn9HhSDVb7dVVKAzHb8pcWPYkyfFd3xt/view?usp=sharing)
* [cell information](https://drive.google.com/file/d/1gfJQEb-ybKFHDeJebtWqHTcjUwr-uFiR/view?usp=sharing)
* [peak count file](https://drive.google.com/file/d/11_yKVraFC4vvMP0b8bJYpbqWHqh9kTZq/view?usp=sharing)

Store them locally in a directory.

In genomics, data integration generally refers to combining different types of biological data to create a more comprehensive understanding of complex biological systems. Integrating multiple data types—such as genomic, transcriptomic, and epigenomic data—allows researchers to capture diverse molecular layers that influence cellular function, uncovering relationships and regulatory mechanisms that might not be evident from a single dataset.

Specifically, in the context of ATAC-seq (chromatin accessibility data) and RNA-seq (transcriptomic data), integration seeks to connect chromatin accessibility (which regions of DNA are open and potentially active) with gene expression (which genes are being transcribed and expressed as mRNA). This integration helps to build a bridge between gene regulation and gene activity.

ATAC- and RNA integration is a useful approach than can be used for addressing: 
(i) the link between Open Chromatin Regions to gene expression, 
(ii) inferring Gene Regulatory Networks interactions 
(iii) performing cell type annotation 
(iv) analyzing dynamical trajectories.


## Pre-processing workflow

### Installations

For the pipeline, we will use the Signac R library. Please, make sure that
you have installed the following libraries by running the following commands.

**If you have installed them yesterday, you do not need to redo this!**

```{r installation, include=TRUE, eval = FALSE}
install.packages('hdf5r')      ## More instructions about installing 
                                ## https://github.com/hhoeflin/hdf5r
install.packages('ggplot2')
install.packages('patchwork')
install.packages('Seurat')
install.packages('Signac') 


if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("GenomicRanges")
BiocManager::install("AnnotationHub")
```


Then, we load the libraries. 


```{r dependencies, include=TRUE}
library(Signac)
library(Seurat)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
library(hdf5r)
library(AnnotationHub)
```


### Loading data

The input formats for the Signac pipeline might vary. However, there are at 
least to important files which are necessary. 

i) Peaks per cell matrix. Containing the read counts per peak and per cell
as a square matrix. 

ii) Fragment file. A list of all recorded fragment reads including peaks which
are not mapped to peaks. These reads might be necessary for computing some 
QC metrics and further analysis.

Optionally, cell annotations might be also added. 

In the following lines we load the peaks counts matrix, fragment files and 
metadata. In the same way as we did for scRNA-Seq we create a Seurat
object to store the data.


```{r, loading_data,  include=TRUE}
counts <- Read10X_h5("~/Library/CloudStorage/Dropbox/00_MBP_Carl/Teaching/OtherCourses/Freiburg_June_2026/atac_pbmc_1k_nextgem_filtered_peak_bc_matrix.h5")

metadata <- read.csv(
  file = file.path("~/Library/CloudStorage/Dropbox/00_MBP_Carl/Teaching/OtherCourses/Freiburg_June_2026/atac_pbmc_1k_nextgem_singlecell.csv"),
  header = TRUE,
  row.names = 1
)

chrom_assay <- CreateChromatinAssay(
  counts = counts,
  sep = c(":", "-"),
  fragments = file.path("~/Library/CloudStorage/Dropbox/00_MBP_Carl/Teaching/OtherCourses/Freiburg_June_2026/atac_pbmc_1k_nextgem_fragments.tsv.gz"),
  min.cells = 10,
  min.features = 200
)

pbmc <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "peaks",
  meta.data = metadata
)

```

The counts matrix can be accessed in the following path:

```{r, show_counts, include=TRUE}
pbmc[['peaks']]$counts[1:5, 1:5]
```

```{r, show_granges, include=TRUE}
granges(pbmc)
```

In the next step we keep only the peaks which are located in conventional 
chromosomes. 


```{r, filtering_peaks, include=TRUE}
peaks.keep <- seqnames(granges(pbmc)) %in% standardChromosomes(granges(pbmc))
pbmc <- pbmc[as.vector(peaks.keep), ]
```

Next, we annotate the peaks to genomic regions. The AnnotationHub R library 
provides access to genome databales like ENSEMBL and UCSC resources. 

```{r,  include=TRUE}
ah <- AnnotationHub()

# Search for the Ensembl 98 EnsDb for Homo sapiens on AnnotationHub
query(ah, "EnsDb.Hsapiens.v98")
ensdb_v98 <- ah[["AH75011"]]


# extract gene annotations from EnsDb
annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)

# change to UCSC style since the data was mapped to hg38
seqlevels(annotations) <- paste0('chr', seqlevels(annotations))
genome(annotations) <- "hg38"

# add the gene information to the object
Annotation(pbmc) <- annotations
```

## Annotating peaks to genomic regions

```{r}
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

peakAnno <- annotatePeak(pbmc@assays$peaks@ranges, 
                         tssRegion=c(-3000, 3000),
                         TxDb=txdb, annoDb="org.Hs.eg.db")
```
```{r}
plotAnnoPie(peakAnno)
```

Most peaks are in the promoter region of genes, then in intronic regions and in intergenic regions!


## Quality control

In single-cell ATAC-seq (scATAC-seq) data, quality control is essential to ensure meaningful interpretation of chromatin accessibility at the single-cell level. Several metrics are commonly used to assess data quality. 

First, the **nucleosome banding pattern** provides insight into chromatin structure by identifying whether fragments align in characteristic periodic patterns, reflecting nucleosome-bound and nucleosome-free regions. 
This pattern helps distinguish high-quality cells with well-defined chromatin states. 

**TSS enrichment score** measures the accumulation of fragments near transcription start sites (TSS), with high scores indicating open, accessible regions around gene promoters, characteristic of active chromatin. 

Additionally, the **total number of fragments in peaks** is evaluated to confirm that sufficient reads fall within identified accessible regions (peaks), a sign of effective enrichment in regulatory regions. 

Lastly, the **ratio of reads in genomic blacklist regions** is checked; these blacklist regions are typically artifactual and non-informative, so a low proportion of reads mapping to them suggests cleaner, high-quality data with minimal background noise. 


In the following chunk these QC metrics are calculated.


```{r,  qc_metrics_calculation, include=TRUE}
# compute nucleosome signal score per cell
pbmc <- NucleosomeSignal(object = pbmc)

# compute TSS enrichment score per cell
pbmc <- TSSEnrichment(object = pbmc)

# add fraction of reads in peaks
pbmc$pct_reads_in_peaks <- pbmc$peak_region_fragments / pbmc$passed_filters * 100

# add blacklist ratio
pbmc$blacklist_ratio <- FractionCountsInRegion(
  object = pbmc, 
  assay = 'peaks',
  regions = blacklist_hg38_unified
)
```

During quality control it is advisable to use jointly use different
QC metrics in order to decide which cells to retain and what to filter.
In the next plot the total number of counts in cells and signal around
TSS are shown as an scatter density plot.


```{r,  scatter_density_counts_per_peak, include=TRUE}
DensityScatter(pbmc, 
               x = 'nCount_peaks', 
               y = 'TSS.enrichment', 
               log_x = TRUE, 
               quantiles = TRUE)
```


The red lines represent 5, 10, 90 and 95 percentiles. So, for example most of the cells
are in a range of ~ 15k and 45k total reads in peaks (nCount_peaks).


Now, we show a nucleosome band patterning. A healthy good quality ATAC-Seq pattern should show a periodicity like the following one.


```{r,  nucleosome_pattern, include=TRUE}
#pbmc$nucleosome_group <- ifelse(pbmc$nucleosome_signal > 4, 'NS > 4', 'NS < 4')
#FragmentHistogram(object = pbmc, group.by = 'nucleosome_group')
FragmentHistogram(object = pbmc, 
                  group.by = "orig.ident")
```


The following violin plots show the separate distributions of the QC metrics. 


```{r,  include=TRUE, fig.width=15, fig.height=6}
VlnPlot(
  object = pbmc,
  group.by = "orig.ident",
  features = c('nCount_peaks', 'TSS.enrichment', 'blacklist_ratio', 'nucleosome_signal', 'pct_reads_in_peaks'),
  pt.size = 0.1,
  ncol = 5
)
```

We can use this information to filter in/out cells. Filtering usually requires 
visual inspection in order to set up thresholds. Outliers in the distributions
might represent doublets, empty droplets, dying cells or artefactual barcodes. 

```{r,  include=TRUE}
pbmc <- subset(
  x = pbmc,
  subset = nCount_peaks > 9000 &
    nCount_peaks < 100000 &
    pct_reads_in_peaks > 40 &
    blacklist_ratio < 0.01 &
    nucleosome_signal < 4 &
    TSS.enrichment > 4
)
pbmc
```

In the Signac pipeline for scATAC-seq data analysis, RunTFIDF and RunSVD are critical functions used for data normalization and dimensionality reduction. 

RunTFIDF applies a **Term Frequency-Inverse Document Frequency (TF-IDF)** transformation to the cell-by-peak matrix, where each entry reflects how often a particular peak is accessible in a cell, adjusted for how frequently it appears across all cells. This normalization enhances the signal of cell-specific accessibility patterns and reduces the influence of universally accessible regions. Following this, RunSVD (Singular Value Decomposition) reduces the dimensionality of the TF-IDF matrix, extracting key components that capture the variation in chromatin accessibility across cells. This enables clustering and visualization of cells based on chromatin accessibility profiles, facilitating identification of cell types and states within scATAC-seq data.


```{r,  include=TRUE}
pbmc <- RunTFIDF(pbmc)
pbmc <- FindTopFeatures(pbmc, min.cutoff = 'q0')
pbmc <- RunSVD(pbmc)
```

In the process of dimensional reduction, peaks data is transformed into a 
different features/components space (like in Principal Component Analysis PCA). 
In the next plot the correlation between each components of the
dimensional reduction and the sequencing depth. Normally we avoid to drive
any conclusion based in features correlated to the sequencing process since
this might be just an artefactual and for that reason we remove this effect
in the visualization step by removing this dimensions.


```{r,  include=TRUE}
DepthCor(pbmc)
```

Component 1 seems to be very much correlated to the sequencing depth, hence we will discard it.

As in the Gene Expression data we can project our data now in two dimensions
in order to visualize the cells. The next


```{r,  include=TRUE}
pbmc <- RunUMAP(object = pbmc, reduction = 'lsi', dims = 2:30)
pbmc <- FindNeighbors(object = pbmc, reduction = 'lsi', dims = 2:30)
pbmc <- FindClusters(object = pbmc, graph.name ='peaks_snn',verbose = FALSE, algorithm = 3)
DimPlot(object = pbmc, label = TRUE) + NoLegend()
```

## Gene Activity Analysis


In scATAC-seq analysis, gene activity analysis aims to infer gene expression potential from chromatin accessibility data, providing a proxy for transcriptional activity in the absence of direct RNA measurements. Since scATAC-seq profiles accessible regulatory regions, we can estimate the activity of genes by measuring the accessibility of their promoters and nearby regulatory elements, such as enhancers. This is often achieved by aggregating accessibility counts from peaks located within or near a gene body, typically within a defined distance from the transcription start site (TSS). These aggregated signals can be used to build a "gene activity matrix," which resembles an expression matrix in scRNA-seq data and enables comparative analysis. Gene activity analysis is especially valuable in multi-modal or integration workflows, as it allows researchers to connect chromatin accessibility profiles to transcriptional states, facilitating cell type annotation and regulatory network inference in single-cell studies.

The following function calculates gene activities.


```{r,  gene_activity_calculation, include=TRUE}
## This step takes some time to run
gene.activities <- GeneActivity(pbmc)
```

The following step performs a normalization of the calculated gene activities.

```{r,  include=TRUE}
# add the gene activity matrix to the Seurat object as a new assay and normalize it
pbmc[['RNA']] <- CreateAssayObject(counts = gene.activities)
rm(gene.activities)
pbmc <- NormalizeData(
  object = pbmc,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(pbmc$nCount_RNA)
)


```

Then, we can visualize specific gene activity in clusters by inspecting
the UMAP projection.


```{r,  , fig.width=14, fig.height=8, include=TRUE}
DefaultAssay(pbmc) <- 'RNA'

FeaturePlot(
  object = pbmc,
  features = c('MS4A1', 'CD3D', 'LEF1', 'NKG7', 'TREM1', 'LYZ'),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3
)
```




