---
output:
  html_document:
    keep_md: yes
---


```{r, done_in_previous_section, include=FALSE}
library(Seurat)
library(tidyverse)

hbrain_path <- '/Users/cbg-mbp-02/Downloads/data/developing_human_brain_count_mtx.rds'
hbrain_mtx <- readRDS(hbrain_path) 


hbrain_seurat <- CreateSeuratObject(
  counts = hbrain_mtx, 
  project = 'human_dev_brain', 
  assay = 'RNA', 
  min.cells = 1, 
  min.features = 1
)
Idents(hbrain_seurat) <- 'human_dev_brain'

mt_genes <- c(
  "ND1", "ND2", "CO1", "CO2",
  "ATP6", "CYB"
)
mt_genes <- mt_genes[mt_genes %in% rownames(hbrain_seurat)]
hbrain_seurat[["percent.mt"]] <- PercentageFeatureSet(hbrain_seurat, features = mt_genes)

hbrain_seu_filtered <- subset(hbrain_seurat, 
                           nFeature_RNA < 4000 &
                                nCount_RNA < 20000 & 
                                    percent.mt < 5)

hbrain_seu_filtered <- NormalizeData(hbrain_seu_filtered)

hbrain_seu_filtered <- ScaleData(hbrain_seu_filtered)

hbrain_seu_filtered <- FindVariableFeatures(hbrain_seu_filtered, nfeatures = 1000)

hbrain_seu_filtered <- RunPCA(hbrain_seu_filtered, 
                        features = VariableFeatures(hbrain_seu_filtered))

hbrain_seu_filtered <- FindNeighbors(hbrain_seu_filtered)
hbrain_seu_filtered <- FindClusters(hbrain_seu_filtered, 
                            resolution = 0.1, 
                            verbose = FALSE)

hbrain.degs <- FindAllMarkers(hbrain_seu_filtered, 
                            logfc.threshold = 1, 
                            min.pct = 0.05, 
                            min.cells.feature = 10, 
                            verbose = FALSE)

hbrain_seu_filtered <- RunUMAP(hbrain_seu_filtered, 
                       dims = 1:7, 
                       verbose = FALSE)

library(openxlsx)
library(HGNChelper)
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_wrapper.R") 
hbrain_seu_filtered <- run_sctype(hbrain_seu_filtered, 
                     assay = "RNA", 
                     scaled = TRUE, 
                     known_tissue_type="Brain",
                     custom_marker_file="https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_short.xlsx", 
                     name="sctype_classification")
```



# Pseudotime analysis of neuronal maturation using Slingshot

In this section, we will perform a pseudotime analysis to study the
transition between immature and mature neurons in the developing
human brain dataset.

Pseudotime methods attempt to reconstruct dynamic biological
processes from static single-cell snapshots. In this example,
we use the `slingshot` package to infer developmental trajectories
between neuronal states.

## Load required libraries


```{r, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(slingshot)
library(SingleCellExperiment)
library(RColorBrewer)
```

## Subset neuronal populations

We focus only on cells annotated as:

* Immature neurons
* Mature neurons

```{r, message=FALSE, warning=FALSE}
neuronal_seu <- subset(
  hbrain_seu_filtered,
  subset = sctype_classification %in% c(
    #"Immature neurons",
    "Mature neurons"
  )
)

table(neuronal_seu$sctype_classification)
```

## Recompute dimensionality reduction

Although the full dataset already contains PCA and UMAP embeddings,
it is good practice to recompute these embeddings after subsetting.

This allows the reduced dimensions to better capture local variation
within neuronal populations.

```{r, message=FALSE, warning=FALSE}
neuronal_seu <- NormalizeData(neuronal_seu)

neuronal_seu <- FindVariableFeatures(
  neuronal_seu,
  nfeatures = 1000
)

neuronal_seu <- ScaleData(neuronal_seu)

neuronal_seu <- RunPCA(
  neuronal_seu,
  features = VariableFeatures(neuronal_seu)
)

neuronal_seu <- RunUMAP(
  neuronal_seu,
  dims = 1:10,
  verbose = FALSE, 
  n.neighbors = 100, 
  min.dist = 0.1
)

neuronal_seu <- FindNeighbors(
  neuronal_seu,
  dims = 1:10, 
  k.param = 50
)

neuronal_seu <- FindClusters(
  neuronal_seu,
  resolution = 0.01,
  verbose = FALSE
)
```

## Visualize neuronal populations

```{r, message=FALSE, warning=FALSE}
DimPlot(
  neuronal_seu,
  reduction = "umap",
  group.by = "sctype_classification",
  pt.size = 1
)
```

## Convert Seurat object to SingleCellExperiment

`slingshot` works with `SingleCellExperiment` objects.

```{r, message=FALSE, warning=FALSE}
sce <- as.SingleCellExperiment(neuronal_seu)
```

## Run Slingshot trajectory inference

We use:

* UMAP coordinates as reduced dimensions
* Seurat clusters as trajectory clusters

We specify the immature neuron cluster as the starting point.

```{r, message=FALSE, warning=FALSE}
table(neuronal_seu$seurat_clusters,
      neuronal_seu$sctype_classification)
```

Inspect the table above and identify which cluster is enriched
for immature neurons.

Replace `"0"` below if necessary.

```{r, message=FALSE, warning=FALSE}
sce <- slingshot(
  sce,
  clusterLabels = "seurat_clusters",
  reducedDim = "UMAP",
  start.clus = "0"
)
```

## Visualize pseudotime trajectories

```{r, message=FALSE, warning=FALSE}
umap <- reducedDims(sce)$UMAP

plot(
  umap,
  col = brewer.pal(9, "Set1")[
    as.numeric(as.factor(colData(sce)$seurat_clusters))
  ],
  pch = 16,
  asp = 1,
  main = "Slingshot trajectory"
)

lines(
  SlingshotDataSet(sce),
  lwd = 3,
  col = "black"
)
```

## Extract pseudotime values

```{r, message=FALSE, warning=FALSE}
pseudotime <- slingPseudotime(sce)

head(pseudotime)
```

We will store the first lineage pseudotime values inside the
Seurat object metadata.

```{r, message=FALSE, warning=FALSE}
neuronal_seu$pseudotime <- pseudotime[,1]
```

## Visualize pseudotime on UMAP

```{r, message=FALSE, warning=FALSE}
FeaturePlot(
  neuronal_seu,
  features = "pseudotime"
)
```

Cells with low pseudotime values are predicted to represent
earlier developmental states, while higher values correspond
to more mature neuronal states.

## Identify genes associated with pseudotime

We calculate correlations between gene expression and pseudotime.

```{r, message=FALSE, warning=FALSE}
expr_matrix <- GetAssayData(
  neuronal_seu,
  layer = "data"
)

pt <- neuronal_seu$pseudotime

valid_cells <- !is.na(pt)

expr_matrix <- expr_matrix[, valid_cells]
pt <- pt[valid_cells]
```

Compute correlations:

```{r, warning=FALSE, message=FALSE}
invisible({
  gene_correlations <- apply(
  expr_matrix,
  1,
  function(x) cor(x, pt, method = "spearman")
)
})


gene_correlations <- sort(
  gene_correlations,
  decreasing = TRUE
)

head(gene_correlations, 20)
```

Genes positively correlated with pseudotime tend to increase
during neuronal maturation.

Genes negatively correlated with pseudotime tend to be associated
with earlier developmental states.

## Visualize top pseudotime-associated genes

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
top_genes <- names(gene_correlations)[1:9]

FeaturePlot(
  neuronal_seu,
  features = top_genes,
  ncol = 3
)
```


[Previous Chapter (Profiling cells)](./07-Profiling_cells.md)|

